Financialisation of Commodity Markets \\\large{Co-movement Behind-the-Scenes}

### 2024-09-26 paper.tex
Deploying to gh-pages from @ bautheac/financialisation-co-movement@64fa6fe96b1e1f8608bf2d0a2d12598e80634503 🚀

Commodity index swaps, commodity index funds (CIFs), commodity-based
exchange traded funds (ETFs) and notes (ETNs) or commodity linked notes
(CLNs) were amongst the main investment vehicles that allowed
institutional and retail investors to build up commodity exposure in the
early 2000s
\citep{boons_stock_2012, henderson_new_2015, irwin_index_2011, unctad_global_2009, schumann_hunger_2011}.
Most of these products were primarily designed to track prominent
commodity indexes among which the S\&P-GSCI, the most popular.\\
Most of its constituents at the time of writing form the basis of the
broad cross-section of commodities that we consider in this study:
corn-\#2 yellow (XCBT), oats (XCBT), soybean meal (XCBT), soybean oil
(XCBT), soybeans (XCBT), wheat-SRW (XCBT), cattle-feeder (XCME),
cattle-live (XCME), lean hogs (XCME), cocoa (IFUS), coffee-C (IFUS),
cotton-\#2 (IFUS), lumber (XCME), orange juice (IFUS), sugar-\#11
(IFUS), natural gas (XNYM), crude oil-WTI (XNYM), gasoline (XNYM),
heating oil (XNYM), copper (XCEC), gold (XCEC), palladium (XNYM),
platinum (XNYM) and silver (XCEC). Unleaded gasoline, the NYMEX legacy
contract for gasoline, stopped trading in 2006 when it was replaced by
Reformulated Gasoline Blendstock for Oxygen Blending (RBOB) with the two
futures series trading alongside for most of the year. For our NYMEX
gasoline data series, we consider unleaded gasoline up to September
1\textsuperscript{st}, 2006 and RBOB gasoline thereafter, date at which
liquidity for the latter overtook that for the former.\\
We also consider the UK metal complex with aluminium (XLME), copper
(XLME), lead (XLME), nickel-primary (XLME), tin-refined (XLME) and zinc
(XLME).

@@

The commodity futures trading commission (CFTC) operates a comprehensive
system of collecting information on market participants known as the
large trader reporting program (LTRP) where it records and reports to
the public position data on market participants who have position levels
in excess of a particular market specific threshold\footnote{Current
  reporting level thresholds available at:
  \href{https://www.ecfr.gov/cgi-bin/retrieveECFR?gp=&SID=970471b8455f4bab7db4110cfde50731&mc=true&r=SECTION&n=se17.1.15_103}{www.ecfr.gov}.}.
The commission collects market data and position information daily from
clearing members, futures commission merchants (FCMs) and foreign
brokers and publishes corresponding summaries weekly in a series of
market specific reports. In these reports, individual traders are
categorised according to the nature of their trading activity on the
basis of self-reported information\footnote{CFTC form 40. Available at:
  \href{https://www.cftc.gov/sites/default/files/idc/groups/public/@forms/documents/file/cftcform40.pdf}{www.cftc.gov}.}
that is subject to review by CFTC staff for reasonableness. Position
information is then aggregated by category and asset type with each
report providing an aggregation for a set of report specific categories
and, for each category, a breakdown between futures only positions and
positions in futures and options combined.\\
In its legacy report, the commitment of trader report (COT) with data
dating back to 1962, the CFTC distinguishes ``commercial'' and
``non-commercial'' market participants where a ``commercial''
participant is defined as one ``{[}\ldots{]} engaged in business
activities hedged by the use of futures and option markets''\footnote{See
  \href{https://www.gpo.gov/fdsys/pkg/CFR-1998-title17-vol1/xml/CFR-1998-title17-vol1-sec1-3.xml}{CFTC
  regulation 1.3, 17 CFR 1.3(z)} for details.}. A third category,
``non-reportable'', aggregates positions for participants not meeting
the reporting threshold. In response to concerns related to category
accuracy\footnote{See for example \citet{ederington_who_2002}.} the CFTC
now refines its classification in the disaggregated commitment of trader
report (DCOT) where participant categories include
``Producer/Merchant/Processor/User'', ``Swap dealer'', ``Money manager''
and ``Other reportable'' as well as in the supplemental commitment of
trader report (SCOT) that details commodity index trader aggregate
positions for thirteen agricultural commodity futures markets. Data for
DCOT and SCOT date back to 2006 and 2007 respectively.\\
We study the 1997/2018 period where we observe futures only position
data from the COT report, the only CFTC report with data available for
the whole period of interest; as well as futures term structure price
and open interest data from Bloomberg.

@@

@@

By grouping together assets by shared attribute dynamics in common
portfolios, asset pricing techniques--and factor models in
particular--are well suited to study co-movement
\citep{fama_common_1993, carhart_persistence_1997, asness_devil_2013, fama_five_factor_2015, hou_digesting_2015, asness_fact_2015, frazzini_betting_2014, asness_quality_2019}.
Careful, theory-based, attribute selection further enables a refined
analysis of market fundamentals
\citep{schwartz_short_2000, miffre_momentum_2007, gorton_fundamentals_2012, cortazar_commodity_2013, yang_investment_2013, daskalaki_factors_2014, szymanowska_anatomy_2014, fernandez_skewness_2018, bakshi_understanding_2019, boons_basis_2019, sakkas_factor_2020}.
We therefore develop a futures-based asset pricing framework that
includes factors constructed using both theoretical and liquidity
considerations and use this framework to carry out our analysis over
four periods of interest. The first, the pre-financialisation period,
starts in 1997 and is naturally followed by financialisation, spanning
the 2004/2008 period, and the financial crisis and its aftermath,
spanning the 2008/2013 period. The last period spans the 2014/2018
period starting with the beginning of the tapering/end of the US
Quantitative Easing program (QE).\\
Using the weekly price, open interest and market position data described
above we construct market portfolios of individual commodity nearby
futures as well as mimicking portfolios for market, commercial hedging
pressure, term structure and liquidity risk factors in returns. While
our market portfolios are long only, equally weighted portfolios of
particular commodity sets, our other mimicking portfolios for risk
factors are constructed as combinations of two sub-portfolios (legs),
one held long, one held short, which respective constituents are pooled
on a weekly basis according to risk factor specific criteria.\\
For each leg, the position weight for a particular asset for a
particular period is contingent on the value of the corresponding
sorting variable and its rank relative to the other assets in the pool.
The sorting variable values for the asset pool are first rescaled so
that the highest (lowest) value becomes 1 (0). Weights are then
determined pairwise, starting from the extremes with the highest and
lowest figures. A coefficient \(\alpha\) is determined by dividing their
difference (1 in this case) by the mean of the rescaled values. The high
figure is attributed a score of \(\frac{1}{n} \times \alpha\) where
\(n\) is the number of assets in the pool while the score for the low
figure is \(\frac{1/n}{\alpha}\). The operation is repeated for the pair
of assets showing the second highest and second lowest values for the
sorting variable and so on until all the assets in the pool are
attributed a score. The individual asset weights are then determined by
dividing each score by the sum of the scores.\\
The period return for a particular mimicking portfolio is calculated as
the difference between the weighted average return for constituents of
the long portfolio and that for the constituents of the short portfolio.
i.e, return on the mimicking portfolio for risk factor \(i\) (\(f_{i}\))
for period \(t\):

\[r_{t}^{f_{i}}=\sum_{j=1}^{x}w^{long}_{j,t} \times r_{j,t}-\sum_{j=n-x}^{n}w^{short}_{j,t} \times r_{j,t}\]
\(n\equiv\) number of commodities in the set considered for mimicking
portfolio construction.\\
\(x = n \cdot s\).\\
\(s\equiv\) selection threshold (\(\frac{1}{3}\) here).\\
\(w^{*}_{j,t}\equiv\) period \(t\) weight for commodity \(j\) on the
long or short leg.\\
\(r_{j,t}\equiv\) period \(t\) front contract return for commodity
\(j\).\\
\(j\equiv\) commodity rank in the ordered set \(Y_{t}^{f_{i}}\).\\
\(Y_{t}^{f_{i}}\equiv\) period \(t\) ordered set of the commodities
considered for mimicking portfolio construction where the ordering rule
is specific to \(f_{i}\).

@@

We use a time series regression approach where we start with regressions
of the US traded individual commodity weekly front contract returns on
returns to the mimicking portfolio for a particular risk factor where
the portfolio is constructed using the whole cross-section of US traded
commodities. The corresponding \(R^{2}\)s are averaged and the process
is repeated for each of our four risk factors (market, hedging pressure,
term structure and liquidity) and periods (1997/2003, 2004/2008,
2008/2013 and 2014/2018). For the long-short factors the results are
reported independently for models where the regressor is the factor
itself or one of its legs (long vs.~short). The same operation is
repeated for the two (\(p=2\)) and three (\(p=3\)) factor model cases,
where regressors include in turn all combinations of any two and three
risk factors respectively, as well as for the four (\(p=4\)) factor
model case where all the risk factor mimicking portfolios are included.
Results are reported for the one-factor model case with the other
results available upon request.

@@

For each period we also implement the analysis described above over
periods of low and high aggregate CHP (\textbf{CHP}), a notion that we
believe is particularly relevant in the context of financialisation
where issues of co-movement and aggregate market behaviour play a
central role. We define \textbf{CHP} as the cross-sectional average of
individual CHPs for a particular pool of assets. i.e., period \(t\)
\textbf{CHP} for asset pool \(x\):

\[\mathbf{CHP}_{t}^{x}=\frac{1}{n}\sum_{j=1}^{n}CHP_{j,t}\] \(n\equiv\)
number of commodities in set \(x\).\\
\(CHP_{i,t}\equiv\) period \(t\) CHP for commodity \(i\) as defined
above.

We further define aggregate backwardation (\textbf{backwardation}) and
contango (\textbf{contango}) \textbf{CHP} regimes as, for a given
period, \textbf{CHP} levels below and above the period's median
respectively and, in contrast with \citep{hong_what_2012} who aggregate
across sub-sectors, the asset pool we consider for \textbf{CHP}
construction include the whole cross-section of US traded commodities.

In the next part of the analysis, our mimicking portfolios for risk
factors, constructed from the set of all US traded commodities are used
as commodity picking devices on the latter commodity set. For each risk
factor and period, individual US traded commodity front returns are
regressed on returns for the corresponding mimicking portfolio.
Individual commodities are sorted by \(R^{2}\) accordingly from highest
to lowest and the top third (highest \(R^{2}\)s) are selected as
commodity picks for the risk factor for the period. i.e, set of picks
\(P_{t}^{f_{i}}\) for risk factor \(i\) (\(f_{i}\)) over period \(t\):

\[P_{t}^{f_{i}}\equiv\left \{ P_{1}, P_{2}, ..., P_{x} \right \}\]
\(x = n \cdot s\).\\
\(n\equiv\) number of commodities in the set considered for the picking
exercise (twenty-four here).\\
\(s\equiv\) selection threshold (\(\frac{1}{3}\) here).\\
\(P_{j}\equiv\) commodity pick \(j\) for risk factor \(f_{i}\) over
period \(t\)\\
\(j\equiv\) rank in the ordered set \(Y_{t}^{f_{i}}\)\\
\[Y_{t}^{f_{i}}\equiv\left \{ \left \{ \tilde{R_{1}^{2}}, \tilde{R_{2}^{2}}, ..., \tilde{R_{n}^{2}} \right \}, \geq \right \}\]
\(n\equiv\) number of commodities in the set considered for picking
(twenty-four here).\\
\(\tilde{R_{j}^{2}}\equiv\) risk factor \(f_{i}\) returns' explanatory
power on commodity \(j\) nearby futures returns over period \(t\) as
defined above.

For each period we report the proportion of time each pick is held in
the long and (/or) short leg of the corresponding risk factor mimicking
portfolio as well as the average pairwise correlation amongst every set
of picks.

In the last part of the analysis, for each period our mimicking
portfolios for risk factors in returns are constructed using each set of
commodity picks independently as the asset pool for portfolio
construction. A similar time series approach to that described above is
then implemented using these newly formed mimicking portfolios. In this
case, a first set of dependent variables includes the sets of commodity
picks themselves, a second set includes the whole cross-section of US
traded commodities while a third set includes the six UK traded metals
considered in the study.

@@

The descriptive statistics for the commodity series considered in the
study displayed in table Table~\ref{tbl-stats-no-regimes} show a clear
pattern of increase in average return during financialisation with an
equally weighted portfolio of the US traded assets showing a mean
returns of 15.9\% significant at the 5\% level vs.~6.9\% with a 10\%
significance in the pre-financialisation period. The increase is even
more pronounced for an equally weighted portfolio of UK traded assets
(metals), 2.47\% (not significant) to 21.8\% (significant at the 10\%
level), which sets out the global nature of the phenomenon. Most
commodities show a higher mean return while five--Lumber (XCME), Natural
gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary
(XLME)--show a lower figure with five significant figures--Crude oil-WTI
(XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and Tin-refined
(XLME)--against none pre-financialisation. Sugar-\#11 wins the race with
a mean return increasing from -4.6\% to 24.9\% followed by copper with
sizeable increases for both markets: 1.5\% to 28.2\% (XCEC) and 0.5\% to
28.6\% (XLME).\\
The pattern of results for volatilities is similar although the increase
is less pronounced. Both equally weighted portfolios shows an increase,
more pronounced for the UK assets, 10\% to 13.6\% (US) vs.~15.7\% to
26.9\% (UK), while out of the thirty assets, eleven show a drop in
volatility: Oats (XCBT), Lean hogs (XCME), Cocoa (IFUS), Coffee-C
(IFUS), Lumber (XCME), Sugar-\#11 (IFUS), Natural gas (XNYM), Crude
oil-WTI (XNYM), Gasoline (XNYM), Heating oil (XNYM), Palladium (XNYM).
Financialisation seems however to have impacted the commodity complex
beyond these mechanical effects of increased long participation ensuing
from the entry on index type investments as our results suggest next.

@@

The cross-sectional nature of financialisation naturally leads us to
extend the individual hedging pressure paradigm by incorporating
aggregate CHP (\textbf{CHP}), a market wide measure of hedging pressure.
The traditional Keynesian hedging pressure paradigm postulates a
negative relationship between the return on an individual commodity
futures and its hedging pressure. The extended Keynesian hedging
pressure paradigm, in the context of financialisation, should also imply
that returns for individual commodities should be high in periods where
\textbf{CHP} is low and vice et versa. In table
Table~\ref{tbl-stats-regimes} the mean returns are shown over high
(\textbf{contango}) and low (\textbf{backwardation}) \textbf{CHP}
regimes for each time period.\\
Over the first period eighteen of the twenty-four US commodities have
higher mean returns during periods of \textbf{backwardation} as compared
to \textbf{contango}, with the difference being statistically
significant at the 10\% level for two commodities, at the 5\% level for
one and at the 1\% level for another. For the remaining six commodities
the higher mean return during \textbf{contango} is not statistically
significant for any of them. For the six UK commodities, all metals,
five of them had higher mean returns during phases of
\textbf{backwardation} as compared to \textbf{contango} in the first
period, with one of these differences being statistically significant at
the 10\% level while for the remaining asset (Zinc) the higher mean
return during \textbf{contango} is not statistically significant. An
equally weighted portfolio of the twenty-four US commodities had a mean
return of 15.4\% during \textbf{backwardation} phases, statistically
significant at the 1\% level, and a mean return of -1.1\% during
\textbf{contango}, both in the first period, with the difference between
them statistically significant at the 5\% level. The corresponding
figures for an equally weighted portfolio of the UK metals were 4.4\%
and 0.3\% respectively. These results provide broad support for an
aggregate version of the Keynesian hedging pressure hypothesis and
suggest that hedgers in the aggregate were engaged in risk transfer
during this period.\\
Over the second period the pattern essentially reverses. For the
twenty-four US commodities over this period, fourteen have higher mean
returns during phases of \textbf{contango} over \textbf{backwardation}.
Four of the six UK metals achieved higher mean returns over
\textbf{contango} relative to \textbf{backwardation} phases with the
difference being statistically significant for one commodity. The
equally weighted US portfolio had a mean return of 15.7\% over
\textbf{backwardation} against 16.2\% during phases of
\textbf{contango}, both statistically significant at the 10\% level. The
difference was considerably more pronounced for the equally weighted UK
metals portfolio with a mean return of 8\% during \textbf{backwardation}
rising to 35.6\% during \textbf{contango}.

@@

In table , we show the average pairwise returns correlation between each
set of picks in each time period. Following on form the earlier
observations, this exercise gives an indication of the extent to which
attribute co-movement reveals price co-movement, which is particularly
relevant for the long-short factors for which the corresponding
attribute is not directly related to price co-movement.\\
In the first period the average picks pairwise correlation for the
long-short factors ranges from 7.1\% (term structure factor picks) to
12.4\% (open interest factor picks) while the market factor picks show
the highest figure (23.7\%). In the second period the figures for the
long-short factors rise sharply; up to 22.9\% from 8.1\% for the average
CHP factor picks, up to 32.4\% from 12.4\% for the open interest factor
picks and up to 23.4\% from 7.1\% for the term structure factor picks.
This increase is stronger than that for the market factor picks in a
relative sense and is a consequence of the dramatic increase in average
pairwise price correlation amongst the US traded metals which dominate
the average CHP factor set of picks and which presence is stronger
amongst both the open interest and term structure factor sets of picks
over the financialisation period. The higher figures for the open
interest factor are driven by the strong presence of the energy complex
amongst the corresponding picks. The latter exhibits much higher average
pairwise correlations than the other sectors in both periods although
lower in the second relative to the first which reinforces the role
played by the metals in the sharp increase in average pairwise
correlation amongst picks for this factor.\\
With regards to \textbf{CHP} regimes, although not very pronounced, a
pattern is noticeable. For the market, open interest and term structure
factor picks, average pairwise correlations are stronger over phases of
\textbf{contango} in the first period and over phases of
\textbf{backwardation} in the financialisation period, eventually
reverting back to being stronger over \textbf{contango} phases in the
crisis period with the CHP factor sets of picks showing the opposite
pattern. Overall, these results suggest an increase in correlation
between non-price attributes co-movement and price co-movement over the
financialisation period. It is also interesting to note that US traded
metals returns were negatively correlated with the rest of US traded
commodities in the first period and positively in the second.

>>>>

Commodity index swaps, commodity index funds (CIFs), commodity-based
exchange traded funds (ETFs) and notes (ETNs) or commodity linked notes
(CLNs) were amongst the main investment vehicles that allowed
institutional and retail investors to build up commodity exposure in the
early 2000s
\citep{boons_stock_2012, henderson_new_2015, irwin_index_2011, unctad_global_2009, schumann_hunger_2011}.
Most of these products were primarily designed to track prominent
commodity indexes among which the S\&P-GSCI, the most popular.\\
Most of its constituents at the time of writing form the basis of the
broad cross-section of commodities that we consider in this study:
corn-\#2 yellow (XCBT), oats (XCBT), soybean meal (XCBT), soybean oil
(XCBT), soybeans (XCBT), wheat-SRW (XCBT), cattle-feeder (XCME),
cattle-live (XCME), lean hogs (XCME), cocoa (IFUS), coffee-C (IFUS),
cotton-\#2 (IFUS), lumber (XCME), orange juice (IFUS), sugar-\#11
(IFUS), natural gas (XNYM), crude oil-WTI (XNYM), gasoline (XNYM),
heating oil (XNYM), copper (XCEC), gold (XCEC), palladium (XNYM),
platinum (XNYM) and silver (XCEC). Unleaded gasoline, the NYMEX legacy
contract for gasoline, stopped trading in 2006 when it was replaced by
Reformulated Gasoline Blendstock for Oxygen Blending (RBOB) with the two
futures series trading alongside for most of the year. For our NYMEX
gasoline data series, we consider unleaded gasoline up to September
1\textsuperscript{st}, 2006 and RBOB gasoline thereafter, date at which
liquidity for the latter overtook that for the former.\\
We also consider the UK metal complex with aluminium (XLME), copper
(XLME), lead (XLME), nickel-primary (XLME), tin-refined (XLME) and zinc
(XLME). This sample cross-section of assets is further divided into
groups for the purpose of the analysis by country (US, GB), sector
(agriculturals, energy, metals) and by sub-sector (agriculturals:
grains, livestock, softs; energy: gas, petroleum; metals: base,
precious).

@@

The commodity futures trading commission (CFTC) operates a comprehensive
system of collecting information on market participants known as the
large trader reporting program (LTRP) where it records and reports to
the public position data on market participants who have position levels
in excess of a particular market specific threshold\footnote{Current
  reporting level thresholds available at:
  \href{https://www.ecfr.gov/cgi-bin/retrieveECFR?gp=&SID=970471b8455f4bab7db4110cfde50731&mc=true&r=SECTION&n=se17.1.15_103}{www.ecfr.gov}.}.
The commission collects market data and position information daily from
clearing members, futures commission merchants (FCMs) and foreign
brokers and publishes corresponding summaries weekly in a series of
market specific reports. In these reports, individual traders are
categorised according to the nature of their trading activity on the
basis of self-reported information\footnote{CFTC form 40. Available at:
  \href{https://www.cftc.gov/sites/default/files/idc/groups/public/@forms/documents/file/cftcform40.pdf}{www.cftc.gov}.}
that is subject to review by CFTC staff for reasonableness. Position
information is then aggregated by category and asset type with each
report providing an aggregation for a set of report specific categories
and, for each category, a breakdown between futures only positions and
positions in futures and options combined.\\
In its legacy report, the commitment of trader report (COT) with data
dating back to 1962, the CFTC distinguishes ``commercial'' and
``non-commercial'' market participants where a ``commercial''
participant is defined as one ``{[}\ldots{]} engaged in business
activities hedged by the use of futures and option markets''\footnote{See
  \href{https://www.gpo.gov/fdsys/pkg/CFR-1998-title17-vol1/xml/CFR-1998-title17-vol1-sec1-3.xml}{CFTC
  regulation 1.3, 17 CFR 1.3(z)} for details.}. A third category,
``non-reportable'', aggregates positions for participants not meeting
the reporting threshold. In response to concerns related to category
accuracy\footnote{See for example \citet{ederington_who_2002}.} the CFTC
now refines its classification in the disaggregated commitment of trader
report (DCOT) where participant categories include
``Producer/Merchant/Processor/User'', ``Swap dealer'', ``Money manager''
and ``Other reportable'' as well as in the supplemental commitment of
trader report (SCOT) that details commodity index trader aggregate
positions for thirteen agricultural commodity futures markets. Data for
DCOT and SCOT date back to 2006 and 2007 respectively.\\
We examine the period from 1997 to 2018, during which we utilize
futures-only position data from the COT report---the only CFTC report
providing data for the entire duration of interest---as well as futures
term structure price and open interest data sourced from Bloomberg. The
sample period is further divided into four distinct intervals. The
pre-financialisation era spans the 1997-2003 period and is naturally
followed by the financialisation phase running from 2004 to 2008 which
in turn leads into the financial crisis and its aftermath, covering 2008
to 2013. The final interval starts in 2013 with the onset of tapering
and the conclusion of the US Quantitative Easing (QE) program.

@@

For each commodity asset we compute the weekly front futures return and
commercial hedging pressure (CHP) time series with the CHP calculated as
the ratio of the number of long commercial positions to the total number
of commercial positions for the considered asset-week combination. For
each asset-period combination we compute the average return and
corresponding volatility over the entire period as well as over phases
of cross-sectional low and high CHP\footnote{Aggregate CHP or
  \textbf{CHP}: see below for a formal definition and description of the
  phase construction process.}. These statistics are also computed for
equally weighted portfolios of the US and GB traded commodity assets
independently.

For each period, the return series for the individual US-traded assets
are regressed on both the relative change on their respective CHP series
and that on the cross-sectional CHP independently. The resulting
\(R^{2}\) values are then averaged by sector and subsector.
Additionally, pairwise correlations are computed for each combination of
assets across the periods of interest and for each individual year
within the full sample period. These correlation coefficients are
subsequently averaged across the entire cross-section of assets, as well
as by country, sector, and subsector. For each period, the coefficients
are also calculated and averaged across phases of low and high
cross-sectional CHP.

This approach sheds light on the evolving interconnectedness among
commodity assets as the complex transitions from traditional market
dynamics---where commodity producers and speculators had distinct
roles---to a more intricate landscape shaped by the influx of financial
investors.

@@

By grouping together assets by shared attribute dynamics in common
portfolios, asset pricing techniques--and factor models in
particular--are well suited to study co-movement
\citep{fama_common_1993, carhart_persistence_1997, asness_devil_2013, fama_five_factor_2015, hou_digesting_2015, asness_fact_2015, frazzini_betting_2014, asness_quality_2019}.
Careful, theory-based, attribute selection further enables a refined
analysis of market fundamentals
\citep{schwartz_short_2000, miffre_momentum_2007, gorton_fundamentals_2012, cortazar_commodity_2013, yang_investment_2013, daskalaki_factors_2014, szymanowska_anatomy_2014, fernandez_skewness_2018, bakshi_understanding_2019, boons_basis_2019, sakkas_factor_2020}.
We therefore develop a futures-based asset pricing framework that
includes factors constructed using both theoretical and liquidity
considerations. Using the weekly price, open interest and market
position data described above we construct market portfolios of
individual commodity nearby futures as well as mimicking portfolios for
market, commercial hedging pressure, term structure and liquidity risk
factors in returns. While our market portfolios are long only
combinations of particular commodity sets, our other mimicking
portfolios for risk factors are constructed as combinations of two
sub-portfolios (legs), one held long, one held short, which respective
constituents are pooled on a weekly basis according to risk factor
specific criteria.\\
All portfolios are equally weighted and returns for the two-legged
factors are calculated as the difference between the average return for
constituents of the long leg and that for the constituents of the short.
i.e, return on the mimicking portfolio for risk factor \(i\) (\(f_{i}\))
for period \(t\):

\[r_{t}^{f_{i}} = \frac{1}{x} \left ( \sum_{j=1}^{x}r_{j,t} - \sum_{j=n-x}^{n} r_{j,t} \right )\]
\(n\equiv\) number of commodities in the set considered for mimicking
portfolio construction.\\
\(x = n \cdot s\).\\
\(s\equiv\) selection threshold (\(\frac{1}{3}\) here).\\
\(r_{j,t}\equiv\) period \(t\) front contract return for commodity
\(j\).\\
\(j\equiv\) commodity rank in the ordered set \(Y_{t}^{f_{i}}\).\\
\(Y_{t}^{f_{i}}\equiv\) period \(t\) ordered set of the commodities
considered for mimicking portfolio construction where the ordering rule
is specific to \(f_{i}\).

@@

We use a time series regression approach where individual commodity
weekly front futures contract returns are regressed on returns to the
mimicking portfolio for a particular risk factor with the corresponding
\(\beta\)s and \(R^{2}\)s averaged. For the long-short factors the
results are reported independently for models where the regressor is the
factor itself or one of its legs (results available upon request).

For each time period considered in the study, being one of the four
multi-year periods described above or any individual year over the
entire sample time period, we carry out our analysis over the entire
period as well as over phases of low and high aggregate CHP
(\textbf{CHP}) over the period; a notion that we believe is particularly
relevant in the context of financialisation where issues of co-movement
and aggregate market behaviour play a central role. We define
\textbf{CHP} as the cross-sectional average of individual CHPs for a
particular pool of assets. i.e., period \(t\) \textbf{CHP} for asset
pool \(x\): \[\mathbf{CHP}_{t}^{x}=\frac{1}{n}\sum_{j=1}^{n}CHP_{j,t}\]
\(n\equiv\) number of commodities in set \(x\).\\
\(CHP_{i,t}\equiv\) period \(t\) CHP for commodity \(i\) as defined
above.

@@

We further define aggregate backwardation (\textbf{backwardation}) and
contango (\textbf{contango}) \textbf{CHP} regimes as, for a given
period, \textbf{CHP} levels below and above the period's median
respectively and, in contrast with \citep{hong_what_2012} who aggregate
across sub-sectors, our \textbf{CHP} construction draws on the full
cross-section of US-traded assets considered in the study.

@@

Table~\ref{tbl-stats-no-regimes} displays descriptive statistics for the
commodity series considered in the study; the results show a clear
pattern of increase in average return during financialisation with an
equally weighted portfolio of the US traded assets showing a mean
returns of 15.9\% significant at the 5\% level vs.~6.9\% with a 10\%
significance in the pre-financialisation period. The increase is even
more pronounced for an equally weighted portfolio of GB traded assets
(metals), 2.47\% (not significant) to 21.8\% (significant at the 10\%
level), which sets out the global nature of the phenomenon. Most
commodities show a higher mean return while five--Lumber (XCME), Natural
gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary
(XLME)--show a lower figure with five significant figures--Crude oil-WTI
(XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and Tin-refined
(XLME)--against none pre-financialisation. Sugar-\#11 wins the race with
a mean return increasing from -4.6\% to 24.9\% followed by copper with
sizeable increases for both markets: 1.5\% to 28.2\% (XCEC) and 0.5\% to
28.6\% (XLME).\\
The pattern of results for volatilities is similar although the increase
is less pronounced. Both equally weighted portfolios shows an increase,
more pronounced for the GB assets, 10\% to 13.6\% (US) vs.~15.7\% to
26.9\% (GB), while out of the thirty assets, eleven show a drop in
volatility: Oats (XCBT), Lean hogs (XCME), Cocoa (IFUS), Coffee-C
(IFUS), Lumber (XCME), Sugar-\#11 (IFUS), Natural gas (XNYM), Crude
oil-WTI (XNYM), Gasoline (XNYM), Heating oil (XNYM), Palladium (XNYM).
Financialisation seems however to have impacted the commodity complex
beyond these mechanical effects of increased long participation ensuing
from the entry on index type investments as our results suggest next.\\
The cross-sectional nature of financialisation naturally leads us to
extend the individual hedging pressure paradigm by incorporating
aggregate CHP (\textbf{CHP}), a market wide measure of hedging pressure.
The traditional Keynesian hedging pressure paradigm postulates a
negative relationship between the return on an individual commodity
futures and its hedging pressure. The extended Keynesian hedging
pressure paradigm, in the context of financialisation, should also imply
that returns for individual commodities should be high in periods where
\textbf{CHP} is low and vice et versa. In Table~\ref{tbl-stats-regimes}
the mean returns are shown over high (\textbf{contango}) and low
(\textbf{backwardation}) \textbf{CHP} regimes for each time period.\\
Over the first period eighteen of the twenty-four US commodities have
higher mean returns during periods of \textbf{backwardation} as compared
to \textbf{contango}, with the difference being statistically
significant at the 10\% level for two commodities, at the 5\% level for
one and at the 1\% level for another. For the remaining six commodities
the higher mean return during \textbf{contango} is not statistically
significant for any of them. For the six GB commodities, all metals,
five of them had higher mean returns during phases of
\textbf{backwardation} as compared to \textbf{contango} in the first
period, with one of these differences being statistically significant at
the 10\% level while for the remaining asset (Zinc (XLME)) the higher
mean return during \textbf{contango} is not statistically significant.
An equally weighted portfolio of the twenty-four US commodities had a
mean return of 15.4\% during \textbf{backwardation} phases,
statistically significant at the 1\% level, and a mean return of -1.1\%
during \textbf{contango}, both in the first period, with the difference
between them statistically significant at the 5\% level. The
corresponding figures for an equally weighted portfolio of the GB metals
were 4.4\% and 0.3\% respectively. These results provide broad support
for an aggregate version of the Keynesian hedging pressure hypothesis
and suggest that hedgers in the aggregate were engaged in risk transfer
during this period.\\
Over the second period the pattern essentially reverses. For the
twenty-four US commodities over this period, fourteen have higher mean
returns during phases of \textbf{contango} over \textbf{backwardation}.
Four of the six GB metals achieved higher mean returns over
\textbf{contango} relative to \textbf{backwardation} phases with the
difference being statistically significant for one commodity. The
equally weighted US portfolio had a mean return of 15.7\% over
\textbf{backwardation} against 16.2\% during phases of
\textbf{contango}, both statistically significant at the 10\% level. The
difference was considerably more pronounced for the equally weighted GB
metals portfolio with a mean return of 8\% during \textbf{backwardation}
rising to 35.6\% during \textbf{contango}.

@@

The onset of financialisation thus seems to have engendered a change in
the nature of hedger's behaviour, at the aggregate level in particular,
taking it away from the traditional Keynesian view of risk shifting.
This phenomenon has been noted in earlier theoretical and empirical work
\citep{danthine_information_1978, stout_why_1998} in different contexts
from which two competing models of behaviour have emerged.\\
The information arbitrage model
\citep{grossman_impossibility_1980, danthine_information_1978, kyle_informed_1989}
implies that hedgers may often be seeking out counterparties to trade
with rather than being purely passive. This issue is also raised in both
\citet{cheng_financialisation_2014} and \citet{stulz_rethink_1996} who
point out that hedgers may be taking a view on prices just as
speculators do. In fact, by hedging away some of their risk, hedgers are
able to speculate more heavily based on their disagreement against
speculators regarding futures price movement
\citep{simsek_speculation_2013}, which fits into a heterogeneous
expectation theory of speculation\footnote{\citet{hirshleifer_speculation_1975};
  \citet{hirshleifer_reply_1976}; \citet{hirshleifer_theory_1977}}. As
\citet{stout_why_1998} points out, this disagreement-based theory of
trading is one of the main reason the public and the law disapproves
speculation as this form of trading is regarded as
non-productive\footnote{\citet{duffie_challenges_2014} also discusses
  some of the challenges faced by a policy treatment of speculative
  trading motivated by differences in beliefs.}.\\
In the second model hedgers are seen as liquidity providers as in
\citet{kang_tale_2020} with this explanation also consistent with the
nature of financialisation which led to the arrival of long term, long
only investors who require substantial liquidity to roll-over their
positions. Table~\ref{tbl-time-reg-US-CHP} supports this view. We
regress commodity futures front returns on individual commercial hedging
pressure (CHP) as well as aggregate CHP (\textbf{CHP}), our market wide
measure of CHP, in separate models and find that the explanatory power
of CHP at the individual commodity level decreases over the
financialisation period with the US cross-sectional average \(R^{2}\)
falling down to 20.8\% from 25.8\% in the previous period while that of
\textbf{CHP} increases (7.1\% vs 5.4\%). Financialisation therefore
seems to have had the effect of loosening the extent to which futures
risk premiums relate to commodity specific commercial hedging motives
while strengthening that to which they relate to broader dynamics such
as cross-sectional liquidity provision. The pattern is strongest for the
metals sector where the average \(R^{2}\) drops down to 19.1\% from
28.1\% between the two periods at the individual level while it rises
from 8.4\% to 15.7\% at the aggregate level. The symmetry breaks in the
crisis period, with the US cross-sectional average \(R^{2}\) at the
individual level bouncing back up (22.5\% vs 20.8\%) while that at the
aggregate level rises further (14.7\% vs 7.1\%), before eventually
resuming post-crisis with a further rise at the individual level (27.2\%
vs 22.5\%) and a drop at the aggregate level (7.5\% vs 14.7\%)
suggesting a return to legacy commodity specific dynamics.

@@

The entire cross-section sees an increase in average regression
coefficient relative to the market factor over the crisis period with
the strongest progression for the GB traded metals followed by the
agricultural sector (Table~\ref{tbl-regressions-index}). The pricing
performance of the factor dominates the US commodity complex over the
period (Table~\ref{tbl-regressions-factors}), showing a 50\% increase in
mean average \(R^{2}\) to over 31\%, while that for both the CHP and
open interest factor decreases and, although increasing, remains modest
for the term structure factors; the increase is strongest for the GB
traded metal assets with an average \(R^{2}\) of almost 34\% for the
period. Taken together, these results indicate the presence of a
systematic factor across the entire cross-section of the US commodities.
For both the CHP and term structure factors, both legs show an increase
in average \(R^{2}\) and while substantially diverging over the
preceding financialisation period the pricing performance for both legs
seems to converge back to a pattern closer to that observed
pre-financialisation.
